Improving Item-Item Similarity Estimation during Collaborative Filtering

### 2014-03-25 paper/paper.tex
edited paragraphs on TrueSim. Changed names to Rand, Pref

To keep the present work focused, we confine ourselves to memory-based
collaborative filtering (CF) in which we work with a logical user-item matrix
with elements corresponding to ratings given by a particular user to a
particular item. Further, we choose to limit our analysis to item-item
similarity but note that these techniques should be equally effective at
improving user-user similarity estimation. An item is now represented as vector
of the ratings they have received from all users. We now formalize the notion of
item-item similarity. Two alternative models for the generation of user-item
ratings are useful here in arriving at two different definitions of the {\em
true} item-item similarity.

In the first model, we imagine users are randomly presented with items to rate.
Given a sufficient period of time, all users will eventually rate all items such
that all elements of an item's vector correspond to a provided rating. The true
similarity in this model of {\em random} rating, which we denote $TrueSimR$, is
then the vector product of two item vectors. That is, 
\begin{align}
TrueSimR(A, B) = \frac{\sum\limits_{u\in U}
r_{u,A}r_{u,B}}{\sqrt{\sum\limits_{u\in U} r_{u,A}^2}
\sqrt{\sum\limits_{u\in U} r_{u,B}^2}}
\end{align}
where $U$ is the set of all users and $r_{u,I}$ is the rating given by user $u$
to item $I$.

In the second model, we imagine an unlimited supply of users each of whom
chooses to rate a subset of items. The lack of a rating is now meaningful since
it indicates a lack of interest of a particular user for a particular item. This
is captured by the model through a default rating of zero \footnote{We address
the issue of rating bias and the precise meaning of a zero valued rating later
in this report.}. At any point in time, we may estimate the true similarity in
this model of {\em preferential} rating, by computing the cosine similarity of
the two item vectors. Only in the limit of inifinite time does this estimate
converge to the true similarity, which we denote $TrueSimP$.  That is,
\begin{align}
TrueSimP(A, B) = lim_{t\to\infty}\frac{\sum\limits_{u\in U_{AB}}
r_{u,A}r_{u,B}}{\sqrt{\sum\limits_{u\in U_A} r_{u,A}^2}
\sqrt{\sum\limits_{u\in U_B} r_{u,B}^2}}
\end{align}
where $U_A$ is the set of users who have rated $A$, $U_B$ is the set of users
who have rated $B$, and $U_{AB}$ is the set of users who have rated both $A$ and
$B$.

>>>>

To keep the present work focused, we confine ourselves to memory-based
collaborative filtering (CF) in which we work with a logical user-item matrix.
Each element of the matrix corresponds to a rating given by a particular user to a
particular item. Further, we choose to limit our analysis to item-item
similarity but note that these techniques should be equally effective at
improving user-user similarity estimation. In this representaion, an item is
a vector of the ratings it has received from all users. We now formalize the 
notion of {\em true} item-item similarity. Two alternative models for the 
generation of user-item ratings are useful in arriving at two different 
definitions of the {\em true} item-item similarity.

In the first model, we imagine users are randomly presented items to rate. Given 
a sufficient period of time, all users will eventually rate all items, creating
item vectors that contain provided ratings from all users. The true similarity 
in this model of {\em random} rating, denoted $TrueSimRand$, is then the vector 
product of two complete item vectors. That is, 
\begin{align}
TrueSimRand(A, B) = \frac{\sum\limits_{u\in U}
r_{u,A}r_{u,B}}{\sqrt{\sum\limits_{u\in U} r_{u,A}^2}
\sqrt{\sum\limits_{u\in U} r_{u,B}^2}}
\end{align}
where $A$ and $B$ are items, $U$ is the set of all users and $r_{u,I}$ is the 
rating given by user $u$ to item $I$.

In the second model, we imagine an unlimited supply of users each of whom
chooses to rate a subset of items. The lack of a rating is now meaningful since
it indicates a lack of interest of a particular user for a particular item. This
is captured by the model through a default rating of zero \footnote{We address
the issue of rating bias and the precise meaning of a zero valued rating later
in this report.}. At any point in time, we may estimate the true similarity in
this model of {\em preferential} rating, by computing the cosine similarity of
the two item vectors. Only in the limit of inifinite time does this estimate
converge to the true similarity, which we denote $TrueSimPref$.  That is,
\begin{align}
TrueSimPref(A, B) = lim_{t\to\infty}\frac{\sum\limits_{u\in U_{AB}}

r_{u,A}r_{u,B}}{\sqrt{\sum\limits_{u\in U_A} r_{u,A}^2}
\sqrt{\sum\limits_{u\in U_B} r_{u,B}^2}}
\end{align}
where $U_A$ is the set of users who have rated $A$, $U_B$ is the set of users
who have rated $B$, and $U_{AB}$ is the set of users who have rated both $A$ and
$B$.
